---
title: "Exploratory analyses of the IBA data"
author: "Rob Goodsell"
date: '2022-06-06'
output:
  workflowr::wflow_html:
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


# 1. Introduction

The purpose of this site is to document and share some preliminary analyses exploring trends in species richness and community composition from the species tables generated from sequencing of the IBA lysate samples. 

Here we use two data files:

1. The species table generated from sequencing of the lysate samples - 'MBs_12.merged.species.tsv'
2. Some general sample information (location, habitat type etc) - 'IBA_samples_master.csv'

### Load packages & data
```{r , message=FALSE , warning=FALSE}
#packages
library(tidyverse)
library(ggbeeswarm)
library(MetBrewer)
library(lubridate)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(vegan)
library(plotly)
library(RColorBrewer)
library(Rtsne)
library(umap)


#Load data
IBA_species <- read_tsv("data/MBs_12.merged.species.tsv")    # Species table

IBA_locs    <- read.csv2("data/IBA_samples_master.csv")  %>% # Sample lookup
  select(trap_ID, sample_ID , habitat , lysate_ID , biomass_grams , 
         lat = trap_lat , lon = trap_long ,collecting_date) %>% 
  mutate(habitat = str_remove(habitat , "\\?")) %>% 
  mutate(habitat = case_when(habitat == "wind_farm" ~ "Wetland" , 
                             habitat == "Cropland/Grassland" ~ "Grassland" , 
                             habitat == "Urban/Cropland" ~ "Cropland",
                             habitat == "Forest/Grassland" ~ "Grassland" , 
                             TRUE ~ habitat))  %>% 
  mutate(week = week(as_date(collecting_date)),
        month = month(as_date(collecting_date))) %>% droplevels()

```

### House keeping
Before any analysis we need to due some general filtering & data tidying steps. The things we need to filter out are:

1. The spike-in sequences.
2. Any non-arthropod sequences.
3. Any unclassified sequences.
4. Any sequences with low (<20) read numbers.
5. The negative / positive control samples. 

We then join with the species information with sample information.

```{r , message=FALSE , warning=FALSE}

#Spike ins
spike_ins <- c("Shelfordella lateralis", "Gryllus bimaculatus", 
               "Drosophila bicornuta",  "Drosophila serrata", 
               "Gryllodes supplicans", "Gryllus campestris")

arthropods  <-  
         # Filter to arthropods
         # Remove unclassified sequences and spike ins
         filter(IBA_species , Phylum == "Arthropoda",
                !str_detect(Species , "unclassified|_X$|_XX$|_XXX$"),    
                !Species %in% spike_ins) %>%   
         # General tidying
         select(-c("Kingdom" , "Phylum" , "Class" ,
                   "Order" , "Family" , "Genus" , "BOLD_bin")) %>%  
         pivot_longer(-Species , names_to = "lysate_ID" , 
                                values_to = "reads") %>%                   
         mutate(lysate_ID = str_remove_all(lysate_ID , "^FL[0-9]+_")) %>%          
         # Remove low frequency reads & control samples
         filter(reads > 20 , !str_detect(lysate_ID , "[Nn]eg|[Pp]os|air")) %>%                      
         left_join(IBA_locs)


```


### 2. Alpha diversity 
A key question is how does insect diversity change across environmental gradients. We already have the habitat types to hand, and these could impact total insect diversity, so I started here. First though, it's important to consider how habitat is distributed geographically. 

```{r , warnings = FALSE , message=FALSE , fig.cap="Trap locations across different habitat types, point size represent cumulative species richness across full sampling year for a single trap." , fig.width=10 , fig.height = 5}
# get map outline
sweden <- ne_countries(country = "sweden" , scale = "large" , returnclass = "sf") %>% 
          st_transform(crs = 4124)

trap_sr_geo <- arthropods %>% 
  group_by(trap_ID , Species ,habitat , month , lat , lon) %>% 
  summarise(total_reads = sum(reads)) %>%
  ungroup() %>% 
  mutate(pres = 1*(total_reads>0)) %>%  # convert reads to presence absence
  group_by(trap_ID , lat, lon, habitat) %>% 
  summarise(species_richness = sum(pres)) %>%
  ungroup() %>% drop_na(lon , lat) 

ggplot()+
  geom_sf(data =sweden)+
  geom_point(data = trap_sr_geo , aes(lon , lat , size = species_richness , 
                                      fill = species_richness), pch=21 , alpha = 1)+
  scale_size_continuous(range = c(1,2))+
  facet_wrap(~habitat , nrow = 1)+
  theme(legend.position = "none" , axis.text = element_blank())
```

From this figure its clear that habitat could be colinear with climatic or other latitudinal gradients, as habitats have different geographic ranges. For example croplands and grasslands are primarily southerly, whilst forests have a more even distribution across the country.   
  
For the moment, to understand changes in $\alpha$ diversity using presence absence data we only look at raw species ricnhess (total number of unique species) between sites. There are clear trends of species richness across different habitats, with grasslands and croplands having the highest number of species, however all habitats are associated with fairly wide distributions.

```{r , warning=FALSE , message=FALSE , fig.cap="Species richness (aggregated by trap), across habitat types. Each point is the cumulative species richness over the full year sampling period" }
trap_sr <- arthropods %>%
        group_by(trap_ID , Species ,  habitat) %>% 
        summarise(total_reads = sum(reads)) %>%
        ungroup() %>% 
        mutate(pres = 1*(total_reads>0)) %>%  # convert reads to presence absence
        group_by(trap_ID , habitat) %>% 
        summarise(species_richness = sum(pres))

trap_sr %>% drop_na(habitat) %>% 
          ggplot(aes(reorder(habitat , - species_richness) , species_richness))+
          geom_violin(aes(fill=habitat) , alpha = .2 , lwd = 0)+
          geom_beeswarm(aes(fill = habitat) , pch=21 , alpha = .7 , size = 1.2)+
          geom_pointrange(stat="summary" , pch = 21 , fill = "white", size = 1.25, fatten = 1.25)+
          scale_fill_manual(values=met.brewer("Moreau", 7))+
          theme_classic()+
          coord_flip()+
          theme(legend.position = "none" , panel.grid.major.y = element_line( linetype = 2) , 
                axis.text = element_text(size = 10),
                axis.title = element_text(size = 14))+
          labs(y = "Species richness" , x = "Habitat" , fill = "Habitat")

```
    

#### 3. Species richness over time
A key question is how species richness changes over time. The plot below displays the change in species richness over time for each habitat. It's clear that all habitats display the same seasonal trend in total species richness, with the highest diversity in the summer months, but a secondary 'peak' in autumn. 

```{r , echo=FALSE , warning=FALSE , message=FALSE , fig.cap="Species richness over time for each habitat type. Points represent species richness of an individial trap, fitted lines are smooth terms from a gam fitted to species richness trends over time"}

trap_sr_ts <- arthropods %>% 
  group_by(trap_ID , Species ,  habitat , week) %>% 
  summarise(total_reads = sum(reads)) %>%
  ungroup() %>% 
  mutate(pres = 1*(total_reads>0)) %>% 
  group_by(trap_ID , habitat , week) %>% 
  summarise(species_richness = sum(pres))

trap_sr_ts %>% drop_na(habitat) %>% 
  ggplot(aes(week , species_richness , group))+
  geom_point(aes(colour = habitat , group=trap_ID) , alpha = .3 , size = .8)+
  geom_smooth(aes(group=habitat) , colour = "black" , lty =  1, lwd = .6)+
  scale_colour_manual(values=met.brewer("Moreau", 7))+
  facet_wrap(~habitat , nrow = 2)+
  theme_minimal()+
  theme(legend.position = "none"  , strip.text = element_text(size=15) ,
        axis.title = element_text(size = 15))+
  labs(y = "Species richness" , x = "Week of the year")

```

### 4. Community structure
As well as the total species richness of communities, its also important to understand their structure, i.e. are there any trends in species turnover. This application displays the results from  different clustering algorithms applied to the species level data, which were applied to highlight trends in community composition. I applied three different methods to reduce the dimensions of the community data:  
  
  
1. Non-metric multidimensional scaling (NMDS).
2. [tSNE](https://www.jmlr.org/papers/volume9/vandermaaten08a/vandermaaten08a.pdf).
3. [UMAP](https://arxiv.org/abs/1802.03426) 

The results are 3d visualizations of community composition for each lysate sample. You can visualize trends according to different covariate data using the drop down menu. There are clear changes in community composition with time (the week variable), and latitude. Hovering over each point will reveal trap information about the sample. 

<iframe src ="https://robgoodsell.shinyapps.io/community_structure_app/" height=800px width=1000px />


